---
title: "Practical Machine Learning Course Project"
author: "Mark Jerome"
date: "April 16, 2017"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project uses the *Weight Lifting Exercises Dataset* found at http://groupware.les.inf.puc-rio.br/har to study predicting from data from accelerometers on the belt, forearm, arm, and dumbell whether a barbell lift has been performed correctly or incorrectly.  In particular, we will develop a model to predict which of five classes a given observation belongs to.  Class A indicates correct form, and classes B, C, D, and E indicate different incorrect forms.

##Overview

**How I built my model:** I built my model using the following steps:

1. Remove line types from training data not found in testing data (new_window="yes").
2. Remove columns which are all blank or all NA.
3. Separate training data into 80% training subset and 20% testing subset.
3. Train five models--one for each class--using one-versus-all logistic regression and 5-fold cross validation.
4. Use the five models to predict a probability that each observation is in each class.
5. For each observation, use the class with the highest probability as the final prediction.
6. Determine expected accuracy from 20% holdout data.

**How I used cross validation:** 5-fold cross validation was used in training each of the five models to help reduce overfitting of the logistic regression coefficients.

**What I think the expected out-of-sample error is:** I expect the out-of-sample accuracy to be .9094, so the out-of-sample error would be .0906.

**Why I made the choices I did:** My choices were made based on speed of processing, desired accuracy, and simplicity.

#Data Preparation

```{r echo=FALSE}
options(warn=-1)
```
```{r echo=TRUE}
set.seed(1234)
library(caret)
train01 <- read.csv("pml-training.csv")
```
The provided training dataset contains two types of observations: ones with *new_window* set to "yes" and ones with *new_window* set to "no".  Because there are no "yes" observations in the provided test dataset, we will not use them in training our model.

```{r}
train02 <- train01[train01$new_window=="no", ]
```
Once the "yes" observations are deleted, there are a number of columns that are either all blank or all NA.  We will not use these to train our model.

```{r}
rawLabels <- train02$classe
keepColNames <- c("user_name","raw_timestamp_part_1","raw_timestamp_part_2","cvtd_timestamp","new_window","num_window","roll_belt","pitch_belt","yaw_belt","total_accel_belt","gyros_belt_x","gyros_belt_y","gyros_belt_z","accel_belt_x","accel_belt_y","accel_belt_z","magnet_belt_x","magnet_belt_y","magnet_belt_z","roll_arm","pitch_arm","yaw_arm","total_accel_arm","gyros_arm_x","gyros_arm_y","gyros_arm_z","accel_arm_x","accel_arm_y","accel_arm_z","magnet_arm_x","magnet_arm_y","magnet_arm_z","roll_dumbbell","pitch_dumbbell","yaw_dumbbell","total_accel_dumbbell","gyros_dumbbell_x","gyros_dumbbell_y","gyros_dumbbell_z","accel_dumbbell_x","accel_dumbbell_y","accel_dumbbell_z","magnet_dumbbell_x","magnet_dumbbell_y","magnet_dumbbell_z","roll_forearm","pitch_forearm","yaw_forearm","total_accel_forearm","gyros_forearm_x","gyros_forearm_y","gyros_forearm_z","accel_forearm_x","accel_forearm_y","accel_forearm_z","magnet_forearm_x","magnet_forearm_y","magnet_forearm_z")
train03 <- train02[, keepColNames]
```
The provided training dataset will be divided into training and testing partitions.

```{r}
inMyTrain <- createDataPartition(rawLabels, p=.8, list=FALSE)
myTest <- train03[-inMyTrain, ]
testLabels <- rawLabels[-inMyTrain]
myTrain <- train03[inMyTrain, ]
trainLabels <- rawLabels[inMyTrain]
```

##Preprocessing

We will center and scale the training data.

```{r}
ppObj <- preProcess(myTrain, method=c("center", "scale"))
input <- predict(ppObj, newdata=myTrain)
ppObj
```

##Training
Define labels to be used in training each of five one-versus-all (ova) models.  There will be one set of labels with two levels for each class: "A" and "nonA", for example.
```{r}
classes <- c("A", "B", "C", "D", "E")
ovaLabels <- data.frame(matrix(nrow=dim(input)[1], ncol=5))
names(ovaLabels) <- classes
for (label in classes){
     ovaLabels[label] <- ifelse(trainLabels==label, label, paste("non", label, sep=""))
}
```
Train five models, one for each class versus all the others.
```{r}
model <- list()
cvControl <- trainControl(method="cv", number=5)
for (label in classes){
     tempDF <- data.frame(ovaLabels[label], input)
     names(tempDF)[1] <- "y"
     x <- train(y~., data=tempDF, method="glm", trControl=cvControl)
     model <- c(model, list(x))
}
names(model) <- classes
```
Use the models to predict a probability that each observation belongs to each class.
```{r}
probabilities <- data.frame(matrix(nrow=dim(input)[1], ncol=5))
names(probabilities) <- classes
for (label in classes){
     probabilities[label] <- predict(model[label], newdata=input, type="prob")[[1]][1]
}
```
Choose the final prediction for each observation from the class with highest predicted probability.
```{r}
predictions <- apply(probabilities, 1, function(x){names(which.max(x))})
```

##Expected Accuracy

Use the 20% holdout data to test model accuracy.
```{r}
input <- predict(ppObj, newdata=myTest)
probabilities <- data.frame(matrix(nrow=dim(input)[1], ncol=5))
names(probabilities) <- classes
for (label in classes){
  probabilities[label] <- predict(model[label], newdata=input, type="prob")[[1]][1]
}
predictions <- apply(probabilities, 1, function(x){names(which.max(x))})
confusionMatrix(predictions, testLabels)
```

##Bibliography

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. **Qualitative Activity Recognition of Weight Lifting Exercises**. *Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13)*. Stuttgart, Germany: ACM SIGCHI, 2013.

